A note on the instability and degeneracy of deep learning models

### 2016-11-23 writing/note.Rmd
vardeman's note edits

---
title: "A note on the instability and degeneracy of deep learning models"
author: | 
  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
  | Iowa State University
  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
date: "November, 2016"
linestretch: 2
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlaated data structures found in several application areas, there is increasing interest in identifying instability in potential model specifications for such data. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes. As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.
  Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence. In this case, a one component change in the data sequence can provably shift probability by orders of magnitude. Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
---

@@

We consider the behavior, and the potential impropriety, of probability models built to incorporate a sequence of discrete observations with an arbitrarily specified length $N$. Let $(X_1, \dots X_N)$ denote a set of sequence of discrete random variables with a finite sample space, $\mathcal{X}^N$. That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ assumes values in the $N$-fold product space $\mathcal{X}^N$. For any $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$. We assume that the model support of $P_{\boldsymbol \theta_N}$ is equal to the sample space $\mathcal{X}^N$. This framework produces a series $P_{\boldsymbol \theta_N}$ of probability models, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data of any given length $N \geq 1$. The size and structure of such parameters may be generally specified, and one may intuitvely consider Euclidean $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$. We will call this class *Finitely Supported Finite Sequence (FSFS) models*.

Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models). Section \ref{instability-results} establishes formal results regarding the model propriety of FSFS models with regard to stability. A FSFS probability model intuitively exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ may result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$. The concept of instability, introduced in the field statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential 
family models by @schweinberger2011instability. For similar exponential models, particularly in connection to random graphs/networks, @handcock2003assessing considered (mean-based) characterizations for so-called model degeneracy, whereby a probability model places all mass on a small subset of the sample space and produces undesirably low variability in model outcomes. As described by @schweinberger2011instability, model instability and model degeneracy are related by viewing degeneracy as an extreme or limiting form of instability. However, the instability results of @schweinberger2011instability are only intended for a special parametric case of exponential family models. In considering the broad framework of FSFS models, the main results here develop a more general measure of model instability, which is defined to arise when certain maximal probabilities in FSFS models are too extreme relative to the length $N$ of the data sample. In this case, a one component change in the data sequence
may provably shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases. Lastly, Section \ref{implications} emphasizes the implications of our model propriety results and proofs of the main results appear in the Appendix.

@@

Many model families can fall under the umbrella of FSFS models. For illustration, this section presents three specific examples of FSFS models, including models with deep architectures.

@@

For $X_1, \dots, X_N$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model with probability mass function of the form
$$
p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
$$ 
where $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ is a vector of natural parameters, $\boldsymbol g : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, positive integers $k$ and $L$ denote the dimensions of the parameter and natural parameters spaces, respectfully, 
$$
\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
$$
is the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.

Defining $\boldsymbol \theta_N=\lambda_N$ to a subsequence of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model. Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables. For these exponential models, the length of the parameter sequence $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).  This aspect may not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}.  We also note that @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for but with fixed parameter sequences $\theta_N=\lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying length $k$.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [cf. @smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$. The RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
$$
p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
$$
where 
$$
\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta,
$$ 
is the normalizing function, $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ is a matrix with dimension $N_H \times N$ are the parameters in the model. Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters.

@@

Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, is expanding in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to vary as an arbitrary function of $N$ as well; in particular, the number $N_H$ of hideen may also potentially increase with $N$. Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the visible RBM model specification is a FSFS model. This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family) is again a FSFS model class.

@@

**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between layers. The joint probability mass function for a DBM is
$$
p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right], 
$$
for $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}$ where
$$
\psi(\boldsymbol \theta) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
$$ 
is the normalizing functionand parameters in the model are given by $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$. Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.

The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification:
$$
p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
$$

Again similar to the RBM case, the visible DBM model specification is an example of a FSFS model.

**Deep belief network (DBN).** A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers. The difference between the two models is that all but the last stacked model are Bayesian networks, rather than RBMs. Thus for visibles $X_1, \dots, X_N$ with support equal to the sample space $\mathcal{X}^N$, a DBN is also a FSFS model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles. Commonly, as in a logistic belief net [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, satisfying the definition of a FSFS model.

@@

To define or measure instability in a FSFS model, it is useful to imagine the behavior of a data model sequence $P_{\theta_N}$ specified to incorporate more and more observations to reach a sample length $N$. A relevant quantity to this end is a (scaled) extremal log-probability ratio (ELPR) given by 
\begin{align}
\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N). \label{eqn:elpr}
\end{align}

The main idea is that, in formulating a FSFS model for potentially increasing numbers of variables (i.e., $N \rightarrow \infty$), the ratio (\ref{eqn:elpr}) should remain bounded, entailing that the largest probability possible under $P_{\theta_N}$ should maintain a magnitude within the smallest probability allowed under the same model. Specifically, the log of the fold change should grow at mostly linearly with the sample size $N$. This leads to the following definition:

\begin{defn}[S-unstable FSFS]
Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$. A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), it holds that 
\begin{align*}
\label{defn:instab}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.

Note that this definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models (e.g. RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an increasing number of parameters. While this definition differs in form and scope, it does match that in @schweinberger2011instability for the special exponential models (cf. Section \ref{discrete-exponential-family-models}) considered there.

S-unstable FSFS models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ by exactly one component}\right\},
$$
the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following non-asymptotic result.

@@

In other words, in S-unstable FSFS models, all probability in the model formulation with a large number of random variables will mass on mode sets of arbitrarily small size or simply on those few outcomes with the highest probability.

>>>>

---
title: "A note on the instability and degeneracy of deep learning models"
author: | 
  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
  | Iowa State University
  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
date: "November, 2016"
linestretch: 2
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlated data structures found in several application areas, there is increasing interest in predicting/identifying instability. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes. (A sequence of probability models results indexed by $N$ accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$. In this case, a one component change in the data sequence can provably shift probability by orders of magnitude. Also, as a measure of instability becomes more extreme, the resulting probability models are shown  to tend to degeneracy, placing all their probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
---

@@

We consider the behavior, and the potential impropriety, of probability models built to incorporate a sequence of discrete observations with length $N$. Let $(X_1, \dots, X_N)$ denote a set of sequence of discrete random variables with a finite sample space, $\mathcal{X}^N$. That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ takes values in the $N$-fold product space $\mathcal{X}^N$. For each $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$. We assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$. This framework produces a series $P_{\boldsymbol \theta_N}$ of probability models, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data of each length $N \geq 1$. (The size and structure of such parameters are without restriction, and natural cases include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.) We will call this model class *Finitely Supported Finite Sequence (FSFS) models*.

Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models). Section \ref{instability-results} establishes formal results regarding the propriety of FSFS models with regard to stability. A FSFS probability model sequence exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ can result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$. The concept of instability, introduced in the field of statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential 
family models by @schweinberger2011instability. For similar exponential models, particularly in connection to random graphs/networks, @handcock2003assessing considered (mean-based) characterizations for so-called model degeneracy, whereby a probability model places all mass on a small subset of the sample space and produces undesirably low variability in model outcomes. As described by @schweinberger2011instability, model instability and model degeneracy are related by viewing degeneracy as an extreme or limiting form of instability. The instability results of @schweinberger2011instability were developed for the case of discrete exponential family models. The main results here concern a general measure of model instability, appropriate accross the whole FSFS model class. This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$. In this case, a one component change in the data sequence may provably shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases. Lastly, Section \ref{implications} emphasizes the implications of our model propriety results and proofs of the main results appear in the Appendix.

@@

Many model families fall under the umbrella of FSFS models. For illustration, this section presents three specific examples of FSFS models, including models with deep architectures.

@@

For $\boldsymbol X = X_1, \dots, X_N$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
$$
p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
$$ 
for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter and natural parameters spaces respectively, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$, $\boldsymbol g : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics, 
$$
\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
$$
the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.

Defining $\boldsymbol \theta_N=\lambda_N$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models. Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables. For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).  This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}. @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\theta_N=\lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [cf. @smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$. For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ is a matrix with dimension $N_H \times N$, the RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
$$
p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
$$
where 
$$
\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta,
$$ 
is the normalizing function. Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.

@@

Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hidens may also potentially increase with $N$. Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model. This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.

@@

**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between neighboring layers. The joint probability mass function for a DBM is
$$
p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right], 
$$
for $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}$ where
$$
\psi(\boldsymbol \theta) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
$$ 
is the normalizing function and parameters in the model are $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$. Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.

The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as
$$
p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
$$

Again like the RBM case, the visible DBM model specification is an example of a FSFS model.

**Deep belief network (DBN).** A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers. The difference between the DBM and DBN models is that all but the last stacked layer in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs. Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N$, a DBN is also a FSFS model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles. Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, satisfying the definition of a FSFS model.

@@

To define or measure instability in FSFS models, it is useful to consider the behavior of a data model sequence $P_{\theta_N}$. A relevant quantity to this end is a (scaled) extremal log-probability ratio (ELPR)
\begin{align}
\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N). \label{eqn:elpr}
\end{align}

The main idea is that, in formulating FSFS models for potentially increasing numbers of variables (i.e., for $N \rightarrow \infty$), the ratio (\ref{eqn:elpr}) should remain bounded, requiring that the largest probability possible under $P_{\theta_N}$ should maintain a fixed order of magnitude relative to the smallest probability allowed under the same model. Specifically, the log of the ratio should grow at mostly linearly with the sample size $N$. This leads to the following definition.

\begin{defn}[S-unstable FSFS]
\label{defn:instab}
Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$. A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), 
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.

This definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models (e.g. RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an increasing number of parameters. While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf. Section \ref{discrete-exponential-family-models}) considered there.

S-unstable FSFS model seqences are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
$$
the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following (non-asymptotic) result.

@@

In other words, in S-unstable FSFS models, all probability in the model formulation with a large number of random variables will concentrate mass on mode sets of arbitrarily small size or simply on those few outcomes with the highest probability.
